---
title: "Bootstrapping and trait integration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bootstrapping and trait integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6
)
```

```{r setup}
library(traitintegration)
```

## Introduction

Once you've learned the syntax for `brms`, using for analyses of ttrait 
integration is very straightforward. But you don't have to use `brms`. If you're
willing to write your own bootstrap code, you can use functions written to 
process bootstrap samples to do the same analyses as you'd do with the model
output from `brms`. This vignette uses the same data as in 
`vignette("Introduction")`.

## Using bootstrapping to get a sample of correlation matrices

First we retrieve the data from [Helianthome](http://www.helianthome.org/), 
store it in a data frame called `dat`, and retain only the columns 
`Leaf C N ratio`, `SLA`, `Primary branches`, `Stem diameter at flowering`,
`Ligule length`, `Phyllaries length`, and `population_id`. Note: If you load the
`readr` library, you don't need the `readr::` prefix for the `read_csv()`.

```{r}
## retrieve the phenotype matrix
##
phenos <- readr::read_csv("http://www.helianthome.org/rest/population/phenotype_matrix/wild",
                          show_col_types = FALSE)
# ## retrieve the list of populations (for population_id)
# ##
# populations <- readr::read_csv("http://www.helianthome.org/rest/population/list/",
#                           show_col_types = FALSE)
## retrieve individual id and population id
##
index <- readr::read_csv("http://www.helianthome.org/rest/individual/list/",
                          show_col_types = FALSE)
## merge population id into phenotypic data
##
wild <- merge(phenos, index, by = "individual_id") |>
  dplyr::select(`Leaf C N ratio`, SLA, `Primary branches`,
                `Stem diameter at flowering`, `Ligule length`,
                `Phyllaries length`, population_id) |>
  dplyr::rename(Leaf.C.N.ratio = `Leaf C N ratio`,
                Primary.branches = `Primary branches`,
                Stem.diameter.at.flowering = `Stem diameter at flowering`,
                Ligule.length = `Ligule length`,
                Phyllaries.length = `Phyllaries length`) |>
  dplyr::filter(!is.na(Leaf.C.N.ratio) &
                !is.na(SLA) &
                !is.na(Primary.branches) &
                !is.na(Stem.diameter.at.flowering) &
                !is.na(Ligule.length) &
                !is.na(Phyllaries.length)) |>
  dplyr::relocate("population_id")
```

Notice that unlike when we use `brms` we have to explicitly drop individuals with
incomplete data. We also `relocate()` the `population_id` column to the front of 
the data frame for convenience with later processing.

Let's talk through the logic of bootstrapping these data before we present the
code. Recall that the individuals are nested within populations and that we're
interested in estimating the within-population correlation matrix. This means
that we need to center and scale the traits *within populations* after selecting
each bootstrap sample. We also will store the bootstrap sample as a 
three-dimensional array with the first dimension being the number of bootstrap
samples, and the second two being the rows and columns of the correlation
matrics.^[This format matches the format of the posterior correlation matrix 
returned from `brms` after processing with `get_Omega()`.] Notice that we have
to be careful with the scaling step, because scaling can fail for one of two
legitimate reasons.

1. A particular bootstrap sample may include only one individual from that 
population. If so, the standard deviation is 0 and scaling produces NaNs.
2. A particular bootstrap sample may include multiple individuals, but a 
particular trait (or traits) may not have any variation, which produces NAs. 

If either of those conditions occur, we drop that population from the 
bootstrap sample that is used to estimate the within-population correlation
coefficient. One other thing to note: The code below includes 
`use = "complete.obs"` in the call to `cor()`. This probably isn't necessary
given the earlier filtering, but this option ensures that all of the 
correlation matrices are positive definite as is required for them to be
valid.

```{r, eval = FALSE}
scale_within_population <- function(dat, pop) {
  tmp <- subset(dat, population_id == pop)
  tmp <- as.data.frame(scale(tmp[, -1]))
  col_names <- c(colnames(tmp), "population_id")
  tmp <- tibble::add_column(tmp, pop)
  colnames(tmp) <- col_names
  tmp <- tmp |>
    dplyr::relocate(population_id)
  return(tmp)
}

scaling_valid <- function(dat) {
  tmp <- as.matrix(dat[, -1])
  result <- sum(is.nan(tmp))
  result <- result & sum(is.na(tmp))
  return(result == 0)
}

scale_within_populations <- function(dat) {
  pops <- unique(dat$population_id)
  tmp <- scale_within_population(dat, pops[1])
  for (i in 2:length(pops)) {
    scaled <- scale_within_population(dat, pops[i])
    ## don't include populations where scaling fails
    ##
    if (scaling_valid(scaled)) {
      tmp <- tibble::add_row(tmp, scaled)
    }
  }
  return(tmp)
}

correlation_within_populations <- function(dat) {
  tmp <- scale_within_populations(dat)
  R <- cor(tmp[, -1], use = "complete.obs")
  return(R)
}

bootstrap <- function(n_sample, dat) {
  n_traits <- ncol(dat) - 1
  Omega <- array(dim = c(n_sample, n_traits, n_traits),
                 dimnames = list(paste("Sample", 1:n_sample, sep = ""),
                                 colnames(dat)[-1],
                                 colnames(dat)[-1]))
  for (i in 1:n_sample) {
    rows <- sample(1:nrow(dat), nrow(dat), replace = TRUE)
    R <- correlation_within_populations(dat[rows, ])
    Omega[i, , ] <- R
  }
  return(Omega)
}

n_sample <- 1000
boot_Omega <- bootstrap(n_sample,  wild)
```

## Visualizing the posterior correlation matrix

Now that we have our bootstrap samples, we can visualize the correlation matrix
and print the bootstrap mean correlation matrix in the same way as we do with 
posterior samples from `brms`. The only difference is that we append `_boot` to
the call to `plot_posterior_correlation()`. `plot_posterior_correlation_boot()`
provides the same options to adjust the display as `plot_posterior_correlation()`.
The plot below uses the default options

```{r}
post_corr <- plot_posterior_correlation_boot(boot_Omega)
print(round(post_corr$R, 3))
```

## Estimating trait integration

Estimating trait integration is straightforward. With a bootstrap sample
we simply call the `assemble_statistics_boot()` function with the bootstrap
sample of correlation matrices. It is best to store the result in a variable, 
say `results`, so that you can work with the results.

```{r}
results <- assemble_statistics_boot(boot_Omega)
str(results)
```

As you can see, `results` is a data frame in which the columns starting with 
`EV_` are the eigenvalues of the correlation matrix, `Lead_EV` is the lead eigenvalue,
`Var_EV` is the variance of the eigenvalues, and `ent_R` the entropy measure
of trait integration. The rows of the data frame correspond to samples from
the posterior distribution.

You can summarize the results using the `summarize_statistics()` function.^[Notice 
that there's no `_boot` suffix on this call. It's the same function as used in 
analysis of output from a `brms` model.]

```{r}
summarize_statistics(results)
```

Since `results` from a call to `assemble_statistics_boot()` is in the same format
as from a call to `assemble_statistics()`, you can visualize the posterior 
distribution of any column using functions from the `bayesplot` package just as
when analyzing results of an analysis with `brms`. For example, you can visualize the
posterior distributions of the lead eigenvalue, variance of eigenvalues, and entR 
as follows (and as before, you can leave off the `bayesplot::` and ``ggplot2::`
prefixes if you've loaded those libraries):

```{r}
bayesplot::mcmc_areas(results, pars = c("Lead_EV", "Var_EV", "ent_R"),
                      prob = 0.95) +
  ggplot2::labs(x = "Trait integration statistics")
```

Note: You will only be able to use `bayesplot` functions that expect all chains
to be merged, e.g., `mcmc_areas()`, `mcmc_intervals()`, mcmc_hist()`, and `mcmc_dens()`.

## Final note

In case you're interested, getting a posterior sample of 4000 correlation matrices
from `brms` takes about 40 seconds on a 16 inch 2023 MacBook Pro with an Apple M3 
processor and 36 GB of RAM. Getting bootstrap sample of 1000 correlation matrices 
takes about 50 seconds on the same machine.
