---
title: "Using `brms` and `traitintegration`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using `brms` and `traitintegration`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(traitintegration)
```

## Introduction

This vignette provides an overview of how to use `brms` to fit a Bayesian 
multiple response model to traits measured on individuals. The goal in doing so 
is to generate a posterior distribution for the correlation among traits. We 
can then use this posterior distribution in the `traitintegration` package to
estimate the posterior distribution of the maximum eigenvalue, the variance of
eigenvalues, and entR as measures of trait integration. We will use the
wild population data from [Helianthome](http://www.helianthome.org/) as an 
example.

## Fitting a model with `brms`

First we retrieve the data from [Helianthome](http://www.helianthome.org/), 
store it in a data frame called `dat`, and retain only the columns 
`Leaf C N ratio`, `SLA`, `Primary branches`, `Stem diameter at flowering`,
`Ligule length`, `Phyllaries length`, and `population_id`. Note: If you load the
`readr` library, you don't need the `readr::` prefix for the `read_csv()`.

```{r}
## retrieve the phenotype matrix
##
phenos <- readr::read_csv("http://www.helianthome.org/rest/population/phenotype_matrix/wild",
                          show_col_types = FALSE)
# ## retrieve the list of populations (for population_id)
# ##
# populations <- readr::read_csv("http://www.helianthome.org/rest/population/list/",
#                           show_col_types = FALSE)
## retrieve individual id and population id
##
index <- readr::read_csv("http://www.helianthome.org/rest/individual/list/",
                          show_col_types = FALSE)
## merge population id into phenotypic data
##
wild <- merge(phenos, index, by = "individual_id") |>
  dplyr::select(`Leaf C N ratio`, SLA, `Primary branches`,
                `Stem diameter at flowering`, `Ligule length`,
                `Phyllaries length`, population_id) |>
  dplyr::rename(Leaf.C.N.ratio = `Leaf C N ratio`,
                Primary.branches = `Primary branches`,
                Stem.diameter.at.flowering = `Stem diameter at flowering`,
                Ligule.length = `Ligule length`,
                Phyllaries.length = `Phyllaries length`)
```

Now that we have the data, we can fit a model using `brms`. Specifically, we 
will fit a model in which we simply fit a random intercept for each population.
This means that the correlation matrix we estimate is the within-population 
covariance matrix. The lines begininng `wild_` define linear models with a 
random intercept for each of the six traits included in the analysis. They are
combined into a single, multiresponse model using the `+` operator in the
call to `brm`.

The `set_rescor(rescor = TRUE)` argument tells `brms` to estimate correlations 
among the traits. The `refresh = 0` prevents `brms` from telling us how many 
iterations it is completing. The `options(mc.cores = parallel::detectCores())` 
tells brms to use as many cores as you have (up to 4) to run the MCMC chains in
parallel. Note: Just as above, if you load the`brms` library, you don't need the
`brms::` prefix for the calls to `bf()` and `brm()`.

```{r, eval = FALSE}
options(mc.cores= parallel::detectCores())

wild_CN <- brms::bf(Leaf.C.N.ratio ~ (1|population_id))
wild_SLA <- brms::bf(SLA ~ (1|population_id))
wild_branch <- brms::bf(Primary.branches ~ (1|population_id))
wild_stem <- brms::bf(Stem.diameter.at.flowering ~ (1|population_id))
wild_ligule <- brms::bf(Ligule.length ~ (1|population_id))
wild_phyllary <- brms::bf(Phyllaries.length ~ (1|population_id))

mod_wild <- brms::brm(wild_CN + wild_SLA + wild_branch + wild_stem +
                        wild_ligule + wild_phyllary + 
                        brms::set_rescor(rescor = TRUE),
                      data = wild,
                      family = "gaussian", 
                      refresh = 0)
```

`brms` will automatically exclude any individuals with missing data. Now
that the model has finished, we can explore some of the results.

```{r}
summary(mod_wild)
```

The most important results here are that the `Rhat`s are all very close to 1.
That means we have good evidence that the chains have converged. It's also good
to see that both the `Bulk_ESS` and `Tail_ESS` are reasonably high (all above 
6000), and especially that they are high for the `rescor` parameters. Those are
the correlations that we're interested.

## Visualizing the posterior correlation matrix

With the fitted model in hand, visualizing the correlation matrix is 
straightforward. If you store the result of the call to `plot_posterior_correlation()`
in a variable, say `p`, you can print the posterior mean correlation matrix as
shown below.

```{r}
p <- plot_posterior_correlation(mod_wild)
print(p$R)
```

The color of each edge reflects the posterior mean of the correlation coefficient
for the nodes that are being connected, red for negative (< -0.2), blue for 
positive (> 0.2), or white for intermediate (-0.2 to 0.2). If you have correlation
coefficients that are especially large in magnitude, they will display as more 
intense colors of red or blue.

You will probably find the labels at each node a bit long. If you do, you can use the
`labels` argument to give them new names. If you'd also like to see values for
the individual correlation coefficients, you can set the `label_edges` argument
to `TRUE`. Note: storing the result of the call to `plot_posterior_correlation()`
in `p` to suppresses printing of the posterior correlation matrix.

```{r}
p <- plot_posterior_correlation(mod_wild, labels = c("Leaf CN", "SLA", "Branches",
                                                     "Stems", "Ligules", 
                                                     "Phyllaries"),
                           label_edges = TRUE)
```

If you look closely, you'll notice that not all of the edges are labeled. That's
because some of those in the middle would overlap if they were displayed.

You aren't stuck with the color palette used here. You can use any of the 
diverging color palettes in `RColorBrewer`. You can also change the number of
digits displayed by using the `digits` argument.

## Estimating trait integration

Estimating trait integration is straightforward. We simply call the 
`assemble_statistics()` function with the fitted model as the argument. It is
best to store the result in a variable, say `results`, so that you can work
with the results.

```{r}
results <- assemble_statistics(mod_wild)
str(results)
```
As you can see, `results` is a data frame in which the columns starting with 
`EV_` are the eigenvalues of the correlation matrix, `Lead_EV` is the lead eigenvalue,
`Var_EV` is the variance of the eigenvalues, and `ent_R` the entropy measure
of trait integration. The rows of the data frame correspond to samples from
the posterior distribution.

You can summarize the results using the `summarize_statistics()` function.

```{r}
summarize_statistics(results)
```

By default the data frame that is returned contains the posterior mean and the
2.5% and 97.5% quantiles of the statistic and rounds the results to 3 digits, but
those options are easy to change.

```{r}
summarize_statistics(results, probs = c(0.025, 0.05, 0.5, 0.95, 0.975), digits = 2)
```

You can also visualize the posterior distribution of any column using functions
from the `bayesplot` package. For example, you can visualize the
posterior distributions of the lead eigenvalue, variance of eigenvalues, and entR 
as follows (and as before, you can leave off the `bayesplot::` and ``ggplot2::`
prefixes if you've loaded those libraries):

```{r}
bayesplot::mcmc_areas(results[, c("Lead_EV", "Var_EV", "ent_R")])
```

Note: You will only be able to use `bayesplot` functions that expect all chains
to be merged, e.g., `mcmc_areas()`, `mcmc_intervals()`, mcmc_hist()`, and `mcmc_dens()`. 
